Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-28 parallel-tractability-J-v9/response.tex

\begin{quote}
First, in the discussion under Lemma 1, it says $|P^*|$ is polynomial in the size of $P$, which is imprecise, as $|P^*|$ is polynomial in the size of $I$ but not necessarily in the size of $R$. This is not a big issue if one considers NC for data complexity. What is less straightforward and more of concern is the statement that Step~2 costs only one time unit. As Step~2 involves checking the applicability of $B_1, \ldots, B_n \to H$ in $G$ and updating $G$, it depends on the size of $G$ and the size of $G$ is not a constant w.r.t.\ that of $I$. Footnote~4 suggests checking applicability of the rule can be done in constant time via an index. What about updating $G$ and maintaining the index?
\end{quote}

In the paragraph under Lemma~1, we made it clear that $|P^*|$ is polynomial in the size of $I$ for a class of datalog programs.

The statement that Step~2 costs only one time unit is based on the context of one processor. Since one processor
is allocated one rule instance (see the first paragraph of Algorithm~$\mathsf{A}_{bsc}$),
the procedure can be completed in constant time units. We rewrote these sentences by clarifying this.

The constant time of checking applicability by a processor is based on a common assumption when
analyzing the parallel complexity. That is the problem is encoded on a parallel configuration.
Consider materializing datalog programs by Algorithm~$\mathsf{A}_{bsc}$. This assumption allows
us mapping all possible substitutions of the atoms and all rule instances in $P^*$ to a unique memory
location. Since each rule instance is allocated to a processor, there is a one-to-one relation between
processors and rule instances. In this way, a processor can check the applicability of its corresponding
rule instance and access the states of an atom occurring in this rule instance in constant time.
Let $a$, $b$ and $c$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively. 
Since $a, b$ and $c$ depend only on the rule set, they are constants.
We further have that the number of 
constant is at most $|\textbf{I}|a$ and the number of all possible rule instances in $P^*$
is at most $c(|\textbf{I}|a)^b$. 
Thus, the memory space for storing atoms and rule instances is polynomial in the size of $\textbf{I}$. 
It is like that there is a polynomial size of index that maps each atom and rule instance to
the corresponding memory location.

The above assumption is also used for analying the computation of transitive closures and
the problem of boolean matrix multiplication.
In the revision, we gave detailed information about the assumption.

### 2018-01-29 parallel-tractability-J-v9/parallel-tractability-J.tex
added the analysis of the complexity of Algorithm A_{bsc}

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in \texttt{NC} complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

In this work, we focus on datalog rewritable ontology languages.
We propose some algorithms, called \texttt{NC} algorithms, to identify several classes of datalog rewritable
ontologies (called \emph{parallelly tractable classes})
such that materialization over them is parallelly tractable.
We then investigate the parallel tractability of materialization over
two datalog rewritable ontology languages DL-Lite and DHL (\emph{Description Horn Logic}).
We show that materialization over any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ is parallelly tractable.
For DHL, there exist ontologies where materialization can hardly be parallelized.
We propose to restrict the usage of DHL such that the restricted ontologies are parallelly tractable.
We further extend the results to an extension of DHL that also allows \emph{complex role inclusion axioms}.
To verify the practical usability of the above results, we analyze real-world datasets and show
that many ontologies expressed in DHL or its extension belong to
the parallelly tractable classes. Based on the results of parallel tractability,
we further implement an optimized parallel algorithm
and compare it to the state-of-the-art reasoner RDFox. The experimental results show that the
optimization used in our implementation results in a better
performance on parallelly tractable ontologies compared with RDFox.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep \texttt{NC} complexity
\end{keyword}

@@

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class $\texttt{NC}$ \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the \texttt{NC} class, one can give an \emph{\texttt{NC} algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above \texttt{NC} algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

@@

\subsection{The $\texttt{NC}$ Complexity}

The parallel complexity class $\texttt{NC}$, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the $\texttt{NC}$ class
can be solved in poly-logarithmic time on a parallel machine with a polynomial number of processors.
We also say that an $\texttt{NC}$ problem can be solved in \emph{parallel poly-logarithmic time}.
Although the $\texttt{NC}$ complexity is a theoretical analysis tool,
it has been shown that many $\texttt{NC}$ problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the $\texttt{NC}$ problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The \texttt{NC} complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in $\texttt{NC}$.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any class of datalog programs where the rule
set of each datalog program is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless \texttt{P}=\texttt{NC}.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an \texttt{NC} algorithm)}.
In this section, we identify such classes of datalog programs by studying different \texttt{NC} algorithms.

@@

We first give the following definition for a parallel tractable class of datalog programs
where an \texttt{NC} algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs where the
same rule set is shared, we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an \texttt{NC} algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an \texttt{NC} algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our $\texttt{NC}$ algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an \texttt{NC} version
and identify the target \texttt{PTD} class.

@@

In this part, we propose a parallel algorithm
(denoted by $\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.\\

\noindent\texttt{Algorithm $\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Recall that $P^*$ denotes the ground instantiation of $P$,
which consists of all possible ground instantiations of rules in $R$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.\footnote{This might not be practically feasible,
but we focus on a theoretical analysis here.}
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,\footnote{Suppose that each processor
    can use $O(1)$ time units to access the state of ground atoms, i.e., whether this ground
    atom has been added to the materialization graph. This can be implemented by maintaining an
    index of polynomial size.}
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss how Algorithm~$\mathsf{A}_{bsc}$ can be restricted to an \texttt{NC} version.
(\uppercase\expandafter{\romannumeral1}) Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$,
it can be checked that $|P^*|$ is polynomial in the size of $\textbf{I}$. 
This is because that each predicate occurring in $\mathbb{P}$ has a constant arity
and Algorithm~$\mathsf{A}_{bsc}$ does not introduce
new constants. This also means that the number of processors is polynomially bounded.
(\uppercase\expandafter{\romannumeral2}) The computing time of \ref{alg1:addFacts}
and \ref{alg1:halt} occupies constant time units because of parallelism.
(\uppercase\expandafter{\romannumeral3}) The main computation part in Algorithm~$\mathsf{A}_{bsc}$ is the
iteration of \ref{alg1:updateG}. Since each processor handles one rule instance, 
\ref{alg1:updateG} costs one time unit on each processor in theory. 
Assume that there is no delay and other communication problems among all the processors.
We have that the computing time of \ref{alg1:updateG} of the whole algorithm costs one time unit as well.
Further, the whole computing time turns out to be bounded by the number of iterations of \ref{alg1:updateG}. (\uppercase\expandafter{\romannumeral4})
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
The input of $\psi$ is the size of $P$ and the output is a non-negative integer.
Based on (\uppercase\expandafter{\romannumeral1}, \uppercase\expandafter{\romannumeral2},
\uppercase\expandafter{\romannumeral3}, \uppercase\expandafter{\romannumeral4}),
for any datalog program $P$, if we use $\psi(|P|)$ to
bound the number of iterations of \ref{alg1:updateG}, then Algorithm~$\mathsf{A}_{bsc}$
is an \texttt{NC} algorithm, denoted by $\mathsf{A}_{bsc}^{\psi}$ (this symbol also
means that $\psi$ occurs as a configuration for Algorithm~$\mathsf{A}_{bsc}$).

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

@@

It is well known that there is an \texttt{NC} algorithm for computing the
transitive closure \cite{Allender07}.
Based on this result and Algorithm~$\mathsf{OPT}$, we propose an optimized variant of Algorithm~$\mathsf{A}_{bsc}$:\\

\noindent\texttt{Algorithm~$\mathsf{A}_{opt}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$, the algorithm
returns a materialization graph $\mathcal{G}$ of $P$. Initially $\mathcal{G}$ is empty. The following steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg3:addFacts}
\item Compute $S_{\textit{\tiny rch}}$ by performing \ref{rch} in Algorithm~$\mathsf{OPT}$; use an \texttt{NC}
    algorithm to compute the transitive closure $S^*_{\textit{\tiny rch}}$ (see \ref{transClos} in Algorithm~$\mathsf{OPT}$);
    update $\mathcal{G}$ by performing \ref{updateG}  in Algorithm~$\mathsf{OPT}$.\label{alg3:updateG}
\item If no node has been added to $\mathcal{G}$ (in \ref{alg3:updateG}), terminate,
    otherwise iterate \ref{alg3:updateG}. \label{alg3:halt}\hfill$\Box$
\end{enumerate}

@@

We obtain an \texttt{NC} variant of $\mathsf{A}_{opt}$ analogously to the process for Algorithm~$\mathsf{A}_{bsc}$.
It can be checked that an iteration of \ref{alg3:updateG} in $\mathsf{A}_{opt}$ costs poly-logarithmic
time, since the main part is computing $S^*_{\textit{\tiny rch}}$ by an \texttt{NC} algorithm.
Thus, if the number of iterations of \ref{alg3:updateG} is upper-bounded by a poly-logarithmical function,
$\mathsf{A}_{opt}$ is an \texttt{NC} algorithm.
Analogously to $\mathsf{A}_{bsc}^{\psi}$, we use $\mathsf{A}_{opt}^{\psi}$ to denote an \texttt{NC} variant.
Specifically, for any datalog program $P$, the number of iterations of \ref{alg3:updateG} in $\mathsf{A}_{opt}$
is bounded by $\psi(|P|)$, where $\psi$ is a poly-logarithmically bounded function.

@@

In this section, we first discuss that Algorithm~$\mathsf{A}_{opt}$ can hardly work in practice.
In order to make Algorithm~$\mathsf{A}_{opt}$ more practical,
we propose to modify Algorithm~$\mathsf{A}_{opt}$ and give an algorithm variant Algorithm~$\mathsf{A}_{prc}$.
We show that Algorithm~$\mathsf{A}_{prc}$ can also be restricted to an $\texttt{NC}$ version
when materializing DHL$(\circ)$ ontologies that follow both of the simple-concept
and the role-concept restrictions.

>>>>

\begin{abstract}
Materialization is an important reasoning service for applications built
on the ontology languages, such as OWL. With a
rapid growth of semantic data, it is challenging to perform
materialization on large-scale ontologies efficiently.
Current work proposes parallel algorithms and employs computing
platforms of high-performance to make materialization sufficiently efficient and scalable in practice.
However, there do exist ontologies, for which
materialization cannot always be improved by parallelism (we use experiments to show this in Section~\ref{sec:evaluation}).
On the other hand, some well-known large-scale ontologies, such as YAGO, have been shown to have
good performance for parallel reasoning, but they are expressed in ontology languages
that are not parallelly tractable in theory, i.e., the efficiency of reasoning may not be improved
on a parallel implementation.
This motivates us to study the problem of \emph{parallel tractability} of ontology
materialization from a theoretical perspective.
That is, we aim to identify the ontologies for which materialization is
parallelly tractable, i.e., in NC complexity.
The results of the parallel tractability can be further used to optimize
materialization algorithms and to guide ontology engineers in creating parallelly tractable ontologies.

In this work, we focus on datalog rewritable ontology languages.
We propose some algorithms, called NC algorithms, to identify several classes of datalog rewritable
ontologies (called \emph{parallelly tractable classes})
such that materialization over them is parallelly tractable.
We then investigate the parallel tractability of materialization over
two datalog rewritable ontology languages DL-Lite and DHL (\emph{Description Horn Logic}).
We show that materialization over any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_\mathcal{R}$ is parallelly tractable.
For DHL, there exist ontologies where materialization can hardly be parallelized.
We propose to restrict the usage of DHL such that the restricted ontologies are parallelly tractable.
We further extend the results to an extension of DHL that also allows \emph{complex role inclusion axioms}.
To verify the practical usability of the above results, we analyze real-world datasets and show
that many ontologies expressed in DHL or its extension belong to
the parallelly tractable classes. Based on the results of parallel tractability,
we further implement an optimized parallel algorithm
and compare it to the state-of-the-art reasoner RDFox. The experimental results show that the
optimization used in our implementation results in a better
performance on parallelly tractable ontologies compared with RDFox.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep NC complexity
\end{keyword}

@@

According to \cite{MotikNPHO14}, many real large-scale ontologies are essentially expressed in the ontology languages
that can be rewritten into datalog rules. Thus, we focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify the classes of datalog rewritable ontologies such that materialization
over these ontologies is parallelly tractable, i.e., in the parallel complexity class NC \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in the NC class, one can give an \emph{NC algorithm} that handles this problem
in parallel computation \cite{Raymond95}.
An NC algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms,
since they are designed for general datalog programs and has \texttt{PTime}-complete complexity.
Thus, we first give NC algorithms that perform materialization,
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallelly tractable classes})
that can be handled by these NC algorithms. To make the proposed NC algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite to clarify how to apply the above NC algorithms
to study the parallel tractability of ontology materialization.
We show that DL-Lite$_{core}$ and DL-Lite$_\mathcal{R}$ are parallelly tractable.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed NC algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze well-known benchmarks and real-world datasets and show that many real ontologies following the
proposed restrictions belong to the parallelly tractable classes.
We implement a system based on an optimized NC algorithm for DHL materialization.
We compare our system to the state-of-the-art reasoner RDFox.
The experimental results show that the optimization proposed in this paper results in a better performance on parallelly
tractable ontologies compared with RDFox.

The rest of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two NC algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study the parallelly tractable materialization of DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto} respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given NC algorithms.
We analyze real-world datasets and evaluate our implementation in Section~\ref{sec:evaluation}.
We then discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}.
The detailed proofs of the theorems and the lemmas in this paper are given in the appendix.

@@

\subsection{The NC Complexity}

The parallel complexity class NC, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the NC class
can be solved in poly-logarithmic time on a PRAM (parallel, random-access machine) with 
a polynomial number of processors. We also say that an NC problem can be solved
in \emph{parallel poly-logarithmic time}.
Although the NC complexity is a theoretical analysis tool,
it has been shown that many NC problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the NC problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The NC complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in NC.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

@@

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any class of datalog programs where the rule
set of each datalog program is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless P=NC.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an NC algorithm)}.
In this section, we identify such classes of datalog programs by studying different NC algorithms.

@@

We first give the following definition for a parallel tractable class of datalog programs
where an NC algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs where the
same rule set is shared, we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an NC algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an NC algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our NC algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an NC version
and identify the target \texttt{PTD} class.

@@

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze the parallel
complexity. This model allows us giving \emph{the parallel assumption}: 
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{bsc}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,\footnote{Suppose that each processor
    can use $O(1)$ time units to access the state of ground atoms, i.e., whether this ground
    atom has been added to the materialization graph. This can be implemented by maintaining an
    index of polynomial size.}
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupies constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus, 
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

When we say that an algorithm is an NC algorithm. It should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{bsc}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{bsc}$ to be an NC algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|\textbf{I}|)$, 
the computing time of Algorithm~$\mathsf{A}_{bsc}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{bsc}$ is an NC algorithm denoted by
$\mathsf{A}_{bsc}^{\psi}$.

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an NC algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

@@

It is well known that there is an NC algorithm for computing the
transitive closure \cite{Allender07}.
Based on this result and Algorithm~$\mathsf{OPT}$, we propose an optimized variant of Algorithm~$\mathsf{A}_{bsc}$:\\

\noindent\texttt{Algorithm~$\mathsf{A}_{opt}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$, the algorithm
returns a materialization graph $\mathcal{G}$ of $P$. Initially $\mathcal{G}$ is empty. The following steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg3:addFacts}
\item Compute $S_{\textit{\tiny rch}}$ by performing \ref{rch} in Algorithm~$\mathsf{OPT}$; use an NC
    algorithm to compute the transitive closure $S^*_{\textit{\tiny rch}}$ (see \ref{transClos} in Algorithm~$\mathsf{OPT}$);
    update $\mathcal{G}$ by performing \ref{updateG}  in Algorithm~$\mathsf{OPT}$.\label{alg3:updateG}
\item If no node has been added to $\mathcal{G}$ (in \ref{alg3:updateG}), terminate,
    otherwise iterate \ref{alg3:updateG}. \label{alg3:halt}\hfill$\Box$
\end{enumerate}

@@

We obtain an NC variant of $\mathsf{A}_{opt}$ analogously to the process for Algorithm~$\mathsf{A}_{bsc}$.
It can be checked that an iteration of \ref{alg3:updateG} in $\mathsf{A}_{opt}$ costs poly-logarithmic
time, since the main part is computing $S^*_{\textit{\tiny rch}}$ by an NC algorithm.
Thus, if the number of iterations of \ref{alg3:updateG} is upper-bounded by a poly-logarithmical function,
$\mathsf{A}_{opt}$ is an NC algorithm.
Analogously to $\mathsf{A}_{bsc}^{\psi}$, we use $\mathsf{A}_{opt}^{\psi}$ to denote an NC variant.
Specifically, for any datalog program $P$, the number of iterations of \ref{alg3:updateG} in $\mathsf{A}_{opt}$
is bounded by $\psi(|P|)$, where $\psi$ is a poly-logarithmically bounded function.

@@

In this section, we first discuss that Algorithm~$\mathsf{A}_{opt}$ can hardly work in practice.
In order to make Algorithm~$\mathsf{A}_{opt}$ more practical,
we propose to modify Algorithm~$\mathsf{A}_{opt}$ and give an algorithm variant Algorithm~$\mathsf{A}_{prc}$.
We show that Algorithm~$\mathsf{A}_{prc}$ can also be restricted to an NC version
when materializing DHL$(\circ)$ ontologies that follow both of the simple-concept
and the role-concept restrictions.
